---
title: "The Big Mac Index"
date: "[Return to Site](https://duncangates.me)"
output: 
  rmdformats::readthedown: 
    self_contained: true # Other options are downcute, robobook, material, readthedown, html_clean, html_docco, lockdown, https://github.com/juba/rmdformats
    thumbnails: false
    lightbox: true
    gallery: false
    highlight: tango
    number_sections: true
    includes:
      after_body: footer.html
---


```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
library(lares)
library(ggthemes)
library(ggrepel)
library(systemfonts)
library(gganimate)
library(pdftools)
library(ggtext)

## Global options
options(max.print="75")
opts_chunk$set(echo=FALSE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=100)

htmltools::tagList(rmarkdown::html_dependency_font_awesome()) # Needed so fa's in footer will show
theme_set(theme_economist())
```

# Tidy Tuesday


```{r}
big_mac <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-12-22/big-mac.csv')
```

## Looking at the data with Lares

```{r}
df_str(big_mac)
str(big_mac)
```


```{r}
big_mac %>% corr_cross() + 
  geom_col(fill = "steelblue") + 
  theme_economist() +
  ggtitle("Greatest cross correlations in data") +
  theme(axis.text.x = element_blank())
big_mac %>% corr_var(usd_raw) + 
  geom_col() + 
  theme_economist() + 
  scale_fill_manual(values = c("red", "steelblue")) +
  ggtitle("Correlations with raw index relative to US Dollar") +
  theme(axis.text.x = element_blank())
big_mac %>% corr_var(gdp_dollar, logs = T, clean = T, top = 20) + 
  geom_col() + 
  theme_economist() + 
  scale_fill_manual(values = c("red", "steelblue")) +
  ggtitle("Correlations with GDP per person in Dollars") +
  theme(axis.text.x = element_blank())
```

## Plotting the data myself

Looks like Turkey messed up so I filter it out in the next plot, Vietnam and Indonesia have similarly ridiculous numbers so I also filter those in the next plot, after that I decided the best way to do this was grouping the plots since there is no level at which all will be plotted reasonably.

```{r}
big_mac_pivot_local <- big_mac %>% 
  select(-iso_a3, -currency_code) %>%
  dplyr::select(local_price, dollar_ex, dollar_price, date, name) %>%
  pivot_longer(!c(date, name), names_to = "currencies", values_to = "value")

big_mac_pivot_local %>%
  dplyr::mutate(currencies = gsub("dollar_ex", "Exchange Rate", currencies), 
                currencies =  gsub("local_price", "Big Mac Local Price", currencies),
                currencies =  gsub("dollar_price", "Big Mac Dollar Price", currencies)) %>%
  ggplot(aes(date, value, color = currencies)) + 
  geom_line() +
  facet_wrap( ~ name, scales = "free_y") +
  scale_y_continuous(labels = scales::dollar) +
  scale_x_date() +
  scale_color_manual(values = c("steelblue", "red", "black")) +
  labs(title = "Dollar Exchange Rate vs. Local Price for Big Mac", color = "", y = "Price", x = "") +
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = "bottom",
        axis.text.x = element_text(angle = 90))
```


```{r}
big_mac_pivot_notlocal <- big_mac %>% 
  select(-iso_a3, -currency_code) %>%
  dplyr::select(-local_price, -dollar_ex, -dollar_price) %>%
  pivot_longer(!c(date, name), names_to = "currencies", values_to = "value")
ggplot(big_mac_pivot_notlocal, aes(date, value, color = currencies)) + 
  geom_line() +
  facet_wrap( ~ name, scales = "free_y") +
  scale_y_continuous(labels = scales::dollar) +
  scale_x_date() +
  scale_color_economist() +
  labs(title = "Exchange Rates", color = "", y = "Price", x = "") +
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = "bottom",
        axis.text.x = element_text(angle = 90))
```

```{r}
# font_add_google("Avenir Next Condensed")
theme_set(theme_minimal(base_family = "Avenir Next Condensed"))
theme_update(
  axis.title = element_blank(),
  axis.text = element_text(color = "grey40"),
  axis.text.x = element_text(size = 20, margin = margin(t = 5)),
  axis.text.y = element_text(size = 17, margin = margin(r = 5)),
  axis.ticks = element_line(color = "grey91", size = .5),
  axis.ticks.length.x = unit(1.3, "lines"),
  axis.ticks.length.y = unit(.7, "lines"),
  panel.grid = element_blank(),
  plot.margin = margin(20, 40, 20, 40),
  plot.background = element_rect(fill = "grey98", color = "grey98"),
  panel.background = element_rect(fill = "grey98", color = "grey98"),
  plot.title = element_text(color = "grey10", size = 32, face = "bold",
                            margin = margin(t = 15)),
  plot.subtitle = element_markdown(color = "grey30", size = 17, 
                                   lineheight = 1.35,
                                   margin = margin(t = 15, b = 40)),
  plot.title.position = "plot",
  plot.caption.position = "plot",
  plot.caption = element_text(color = "grey30", size = 15,
                              lineheight = 1.2, hjust = 0, 
                              margin = margin(t = 40)),
  legend.position = "none"
)
```

```{r prep-data}
df_mac <- 
  big_mac %>% 
  mutate(year = lubridate::year(date)) %>% 
  dplyr::select(date, year, iso_a3, currency_code, name, dollar_price) %>% 
  group_by(iso_a3, name, year) %>% 
  summarize(price = mean(dollar_price)) %>% 
  group_by(iso_a3) %>% 
  filter(n() == 21)
highlights <- c("EUZ", "CHE", "DNK", "SWE", "BRA", "ARG", "GBR", "USA")
n <- length(highlights)
```

## Animation

```{r prep-index-plots-animation}
index_it <- function(y) {
  countries <- 
    df_mac %>% 
    filter(year == y) %>% 
    pull(iso_a3)
  
  df_mac_indexed <- 
    df_mac %>% 
    group_by(iso_a3) %>%
    filter(iso_a3 %in% countries) %>% 
    mutate(
      ref_year = y,
      price_index = price[which(year == y)],
      price_rel = price - price_index,
      group = if_else(iso_a3 %in% highlights, iso_a3, "other"),
      group = as.factor(group)
    ) %>% 
    mutate(
      group = fct_relevel(group, "other", after = Inf),
      name_lab = if_else(year == 2020, name, NA_character_)
    ) %>% 
    ungroup()
}
df_mac_index_refs <- map_df(2000:2019, ~ index_it(.x))
```


```{r plot-and-animation}
index_plots <- 
  ggplot(df_mac_index_refs %>% filter(group != "other"), 
         aes(year, price_rel, group = iso_a3)) + 
  ## geometric annotations
  geom_vline(
    xintercept = seq(2000, 2020, by = 5),
    color = "grey91", 
    size = .6
  ) +
  geom_segment(
    data = tibble(y = seq(-6, 6, by = 2), x1 = 2000, x2 = 2020),
    aes(x = x1, xend = x2, y = y, yend = y),
    inherit.aes = FALSE,
    color = "grey91",
    size = .6
  ) +
  geom_segment(
    data = tibble(y = 0, x1 = 2000, x2 = 2020),
    aes(x = x1, xend = x2, y = y, yend = y),
    inherit.aes = FALSE,
    color = "grey40",
    size = .5
  ) +
  geom_vline(
    aes(xintercept = ref_year), 
    color = "grey40",
    linetype = "dotted"
  ) +
  ## grey lines
  geom_line(
    data = df_mac_index_refs %>% filter(group == "other"),
    color = "grey70",
    size = .6,
    alpha = .5
  ) +
  ## colored lines
  geom_line(
    aes(color = group),
    size = .9
  ) +
  ## text annotations
  geom_text(
    aes(x = ref_year, y = 5.2, 
        label = as.character(ref_year)),
    stat = "unique",
    family = "Avenir Next Condensed",
    size = 8,
    color = "grey60",
    hjust = 0,
    nudge_x = .3
  ) +
  geom_text_repel(
    aes(color = group,
        label = name_lab),
    family = "Avenir Next Condensed",
    fontface = "bold",
    max.overlaps = 20,
    size = 6.7,
    direction = "y",
    xlim = c(2020.8, NA),
    hjust = 0,
    segment.size = .7,
    segment.alpha = .5,
    segment.linetype = "dotted",
    box.padding = .4,
    segment.curvature = -0.1,
    segment.ncp = 3,
    segment.angle = 20
  ) +
  ## coordinate system + scales
  coord_cartesian(
    clip = "off",
    xlim = c(2000, 2026), 
    ylim = c(-6, 6)
  ) +
  scale_x_continuous(
    expand = c(0, 0),
    breaks = seq(2000, 2020, by = 5)
  ) +
  scale_y_continuous(
    expand = c(0, 0),
    breaks = seq(-6, 6, by = 2),
    labels = glue::glue("{format(seq(-6, 6, by = 2), nsmall = 2)}$")
  ) +
  scale_color_manual(
    values = c(rcartocolor::carto_pal(n = n, name = "Bold")[1:n-1], "grey50")
  ) +
  labs(
    title = "Compared to {closest_state}, how much more or less do you have to pay for a Big Mac today?",
    subtitle = "The animated series of <i>index chart</i> visualizes the price changes (in USD) of a Big Mac based on a selected index year. The <b>Big Mac Index</b> is published by The Economist as an informal<br>way to provide a test of the extent to which market exchange rates result in goods costing the same in different countries. It <i>seeks to make exchange-rate theory a bit more digestible</i><br>and takes its name from the Big Mac, a hamburger sold at McDonald's restaurants.",
    caption = "The index chart shows the 27 countries that provide Big mac prices for all years from 2000 to 2020. In case a country was reported twice per year, the mean value was visualized.\nVisualization by Cédric Scherer  •  Data by The Economist"
  ) +
  ## animation settings
  transition_states(ref_year, 5, 25, wrap = FALSE) +
  ease_aes('quadratic-in-out')
index_anim <-
  animate(index_plots, nframes = 150,
          fps = 5, detail = 5, end_pause = 25,
          width = 1150, height = 740)
anim_save(filename = "2020_52_BigMacIndex.gif", 
          path = here::here("Gifs/2020_52BigMac/"), height = 8, width = 16, units = "in", res = 300)
```


## Static Plot

```{r static-index-plot, fig.width=20, fig.height=12}
countries <- 
  df_mac %>% 
  filter(year == 2008) %>% 
  pull(iso_a3)
  
df_mac_indexed_2008 <- 
  df_mac %>% 
  group_by(iso_a3) %>%
  filter(iso_a3 %in% countries) %>% 
  mutate(
    ref_year = 2008,
    price_index = price[which(year == 2008)],
    price_rel = price - price_index,
    group = if_else(iso_a3 %in% highlights, iso_a3, "other"),
    group = as.factor(group)
  ) %>% 
  mutate(
    group = fct_relevel(group, "other", after = Inf),
    name_lab = if_else(year == 2020, name, NA_character_)
  ) %>% 
  ungroup()
ggplot(df_mac_indexed_2008 %>% filter(group != "other"), 
       aes(year, price_rel, group = iso_a3)) + 
  ## geometric annotations
  geom_vline(
    xintercept = seq(2000, 2020, by = 5),
    color = "grey91", 
    size = .6
  ) +
  geom_segment(
    data = tibble(y = seq(-4, 3, by = 1), x1 = 2000, x2 = 2020),
    aes(x = x1, xend = x2, y = y, yend = y),
    inherit.aes = FALSE,
    color = "grey91",
    size = .6
  ) +
  geom_segment(
    data = tibble(y = 0, x1 = 2000, x2 = 2020),
    aes(x = x1, xend = x2, y = y, yend = y),
    inherit.aes = FALSE,
    color = "grey60",
    size = .8
  ) +
  geom_vline(
    aes(xintercept = ref_year), 
    color = "grey40",
    linetype = "dotted",
    size = .8
  ) +
  ## grey lines
  geom_line(
    data = df_mac_indexed_2008 %>% filter(group == "other"),
    color = "grey75",
    size = .6,
    alpha = .5
  ) +
  ## colored lines
  geom_line(
    aes(color = group),
    size = .9
  ) +
  ## text annotations
  annotate(
    "text", x = 2008.15, y = -3.35, 
    label = "2008",
    family = "Avenir Next Condensed",
    size = 8,
    color = "grey40",
    hjust = 0
  ) +
  geom_text_repel(
    aes(color = group,
        label = name_lab),
    family = "Avenir Next Condensed",
    fontface = "bold",
    size = 8,
    direction = "y",
    xlim = c(2020.8, NA),
    hjust = 0,
    segment.size = .7,
    segment.alpha = .5,
    segment.linetype = "dotted",
    box.padding = .4,
    segment.curvature = -0.1,
    segment.ncp = 3,
    segment.angle = 20
  ) +
  ## coordinate system + scales
  coord_cartesian(
    clip = "off",
    ylim = c(-4, 3)
  ) +
  scale_x_continuous(
    expand = c(0, 0),
    limits = c(2000, 2023.5), 
    breaks = seq(2000, 2020, by = 5)
  ) +
  scale_y_continuous(
    expand = c(0, 0),
    breaks = seq(-4, 3, by = 1),
    labels = glue::glue("{format(seq(-4, 3, by = 1), nsmall = 2)}$")
  ) +
  scale_color_manual(
    values = c(rcartocolor::carto_pal(n = n, name = "Bold")[1:n-1], "grey50")
  ) +
  labs(
    title = "Compared to the financial crisis in 2008, how much more or less do you have to pay for a Big Mac today?",
    subtitle = "The <i>index chart</i> visualizes the price changes (in USD) of a Big Mac based on a 2008 as index year. The <b>Big Mac Index</b> is published by The Economist as an informal way to provide a test of the<br>extent to which market exchange rates result in goods costing the same in different countries. It <i>seeks to make exchange-rate theory a bit more digestible</i> and takes its name from the Big Mac,<br>a hamburger sold at McDonald's restaurants.",
    caption = "Visualization by Cédric Scherer  •  Data by The Economist  •  The index chart shows the 27 countries that provide Big mac prices for all years from 2000 to 2020. In case a country was reported twice per year, the mean value was visualized."
  ) +
  theme(plot.subtitle = element_markdown(size = 20))
```






